Interactive Curvature Tensor Visualization on Digital Surfaces\thanks{This work has been mainly funded by \textsc{DigitalSnow} ANR-11-BS02-009, \textsc{KIDICO} ANR-2010-BLAN-0205 and \textsc{PRIMES Labex} ANR-11-LABX-0063/ ANR-11-IDEX-0007 research grants.}

### 2015-10-22 latex/article.tex
minor edit

\title{Interactive Curvature Tensor Visualization on Digital
Surfaces\thanks{This work has been mainly funded by XXXXXXX research grants.}}

@@

Volumetric objects are being more and more popular in many
applications ranging from object modeling and rendering in Computer
Graphics to geometry processing in Medical Imaging or Material
Sciences. When considering large volumetric data, interactive
visualization of such images (or isovalue surfaces) is a complex
problem. Such issues become even more difficult when dynamic
volumetric datasets are considered. Beside such isosurface
visualization, we are also interested in performing geometry
processing on the digital object and to explore different parameter
settings of the geometry processing tool.  In this work, we focus on
interactive visualization of volumetric data isosurfaces combined with
a curvature tensor (mean/Gaussian curvature, principal curvatures
directions\ldots) estimation at each point of the digital
surface. Curvature maps on a given isosurface are key information when
exploring 3D structures in Material Sciences \todo{REF}. Furthermore,
most curvature estimators require a parameter fixing the scale
to which the computation is performed. For short, such parameter
(integration radius, convolution kernel size, \ldots) specifies noise
level on the object surface. As a consequence, exploring interactively
different values of such parameter is mandatory.
\vspace{0.2cm}

@@

\sloppy\noindent\textbf{Related works}\quad Extracting and Visualizing
isosurface on volumetric data has been widely investigated since the
seminal Marching-Cubes approach by Lorensen and
Cline~\cite{lorensen1987marching}. This approach being data parallel,
GPU implementation of this method is very efficient
\cite{tatarchuk2007real}. However, such technique generates a lot of
triangles which is not well suited to large digital data. Hence,
adaptive approaches have been proposed in order to optimize the
triangular mesh resolution to the object geometry or depending on the
viewpoint. In this topic, many solutions have been developed in
Computer
Graphics~\cite{shu1995adaptive,schaefer2004dual,lengyel2010voxel,DBLP:journals/cgf/LewinerMPPL10,DBLP:journals/cvgip/LobelloDD14}.
The method developed by Lengyel \textit{et al.}
\cite{lengyel2010voxel} is the best suited to our needs.  However, it
was only developed on CPU to visualize in real-time static or near
static data, combining octree with new Marching-Cubes configurations
to generate adaptive meshes. In this paper we propose a GPU pipeline
inspired from this method.  It leaves the CPU free for other tasks and
avoids synchronization with the GPU.  Our high framerate also allows
this method to run in real-time on fully dynamic data.  The
triangulation is controlled by a GPU LoD criterion that ensures that
triangles project onto more than one pixel, thus removing geometrical
aliasing.

@@

In \cite{CVIU2014}, authors have also defined the 3D digital principal curvature
estimators $\PrincCurvH{1}{R}$ and $\PrincCurvH{2}{R}$ on $Z\subset\Z^3$ based
on digital moments:
%
\begin{Definition}
  %
  Given the Gauss digitization $Z \EqDef \DSh$ of a shape $\Shape \subset \R^3$,
  {\em 3D digital principal curvature estimators} are defined for any point $\vx
  \in \R^3$ as:
  %
  \begin{align}
    \forall 0 < h < R,\quad
    \PrincCurvH{1}{R}(Z,\vx,h) &\EqDef \frac{6}{\pi {R}^6}(\hat{\lambda}_2 - 3\hat{\lambda}_1) + \frac{8}{5 R} \,,\label{eq:dig-curvature-estimator-k1}\\
    \PrincCurvH{2}{R}(Z,\vx,h) &\EqDef \frac{6}{\pi {R}^6}(\hat{\lambda}_1 - 3\hat{\lambda}_2) + \frac{8}{5 R} \,,
    \label{eq:dig-curvature-estimator-k2}
  \end{align}
  %
  where $\hat{\lambda}_1$ and $\hat{\lambda}_2$ are the two greatest eigenvalues
  of the covariance matrix of the intersection between the digital shape
  $\DigShape$ and a ball of digital radius $R/h$ centered on digital point
  $\vx/h$.
  %
\end{Definition}
%
The covariance matrix needs to compute digital moments of order $0$
($\DMom{000}{h}$), of order $1$ ($\DMom{100}{h}$, $\DMom{010}{h}$ and
$\DMom{001}{h}$) and of order $2$ ($\DMom{200}{h}$, $\DMom{020}{h}$,
$\DMom{002}{h}$, $\DMom{110}{h}$, $\DMom{011}{h}$ and $\DMom{101}{h}$, see
Equation~20 of \cite{CVIU2014} for more details). They have also proved the
convergence of these estimators when setting the ball radius $h$ in $R =
kh^\frac{1}{3}$, where $k$ is a constant related to the maximal curvature of the
shape, on convex shapes with at least $C^3$-boundary and bounded curvature
\cite{CVIU2014}. More formally, they get:
%
\begin{align}
  \forall 0 < h \le h_0,\,\, &\forall \vx \in \Bd{\Shape},\,\,
  \forall \vxH \in \Bd{\Body{\DSh}{h}} \text{~with~} \| \vxH -\vx\|_\infty \le h, \nonumber \\
  &\PrincCurvH{1}{R}(\DigF{\Shape}{h},\vxH, h) = \PrincCurv{1}(\Shape,\vx) + O\left(h^{\frac{1}{3}}\right) \,,\\
  &\PrincCurvH{2}{R}(\DigF{\Shape}{h},\vxH, h) = \PrincCurv{2}(\Shape,\vx) + O\left(h^{\frac{1}{3}}\right) \,.
\end{align}

Additionally, eigenvectors associated with $\hat{\lambda}_1$ and
$\hat{\lambda}_2$ of the covariance matrix correspond to principal curvature
directions $\PrincDirH{1}{R}$ and $\PrincDirH{2}{R}$. The smallest eigenvector
corresponds to the normal direction $\NormalDirH{R}$ at $\vx$. Convergence
results can be found in \cite{ChapterIICurvature}.\todo[inline]{Changer
ChapterIICurvature dans le bibtex...}

@@

In this section, we detail the adaptive isosurface extraction
algorithm. The proposed approach is based on an octree representation
of the input object on which an efficient adaptive triangulation
of the isosurface is constructed on GPU. Such approach is motivated by
the fact that the hierarchical representation of the object allows us
to handle large datasets and to locally adapt the level of details
with respect to the geometry or camera position. We first present the
octree representation and then the isosurface extraction from
hierarchical octree cells.

@@

\begin{figure}[!htbp]
  \begin{center}
    \subfigure{}{\includegraphics[width=0.9\textwidth]{figs/morton}}
    %%\subfigure{}{\includegraphics[height=0.17\textheight]{figs/partitioning}}
    %%\subfigure{}{\includegraphics[height=0.17\textheight]{figs/quadtree_activefront}}
  \end{center}
\caption{Morton codes associated to cells of a linear quatree. Each
   morton code of a child cell is obtained by adding a suffix to its
  parent code \emph{(left)}. The adaptive representation consists of
  quadtree cells whose depth is view point dependent
  \emph{(middle)}. Finally, adaptive Marching-Cubes is used to
  generate the triangulation \emph{(right)}.}
\label{fig_quadtree_partitionning}
\end{figure}

Representing a hierachical structure on GPU is usually challenging
since the GPU memory is limited and non consistent \todo{fix coherent/aligned/coalesced?}
memory access (\emph{e.g.} tree traversal using pointer dereferencing)
may lead to huge payload from cache misses.  Efficient spatial tree
encoding can be achieved using pointerless structures. We use
linear quadtrees or octrees to represent classical quadtree/octreee
trees, see for instance Gargantini
\cite{gargantini1982effective}. This structure consists in indexing
each cell by a \emph{Morton code}: the code of children cells are
defined by the code of the parent suffixed by two bits (in dimension
2, three bits in dimension 3) (see Figure
\ref{fig_quadtree_partitionning}-\emph{left}). A cell's code encodes
 its position with respect to its parent cell and its
complete path to the tree root. Hence, whatever their depth, all cell
codes are simply stored in a linear vector on memory. Furthermore,
cell operations such as subdivision, merging, neighbors access can be
efficiently implemented using bitwise operations on the Morton code.
In the following, we use the GPU implementation proposed by Dupuy
\textit{et al.}  \cite{dupuy2014quadtrees}. By using their approach, Morton
codes and bitwise operations are optimized to match with GPU hardware
requirements.

@@

Using this spatial datastructure, triangulated mesh can be constructed
from a Marching-Cubes approach \cite{lorensen1987marching} (MC for short): the
triangulation is generated from local triangle patches computed on
local cell configurations. Since triangle patches are locally
computed, such approach is fully parallel and easy to implement on GPU
hardwares. However, since adjacent cells may not have the same depth
on the octree, original Lorensen and Cline's rules need to be
updated (see Figure~\ref{fig_quadtree_partitionning}-\emph{right}). Many authors have addressed this problem both for primal and
dual meshes
\cite{shu1995adaptive,schaefer2004dual,lengyel2010voxel,DBLP:journals/cgf/LewinerMPPL10,DBLP:journals/cvgip/LobelloDD14}.

In the following, we use the extension of MC configurations to handle
adaptive structures proposed by Lengyel \emph{et al.}
\cite{lengyel2010voxel}. First this approach constrains the octree
structure in order to make sure that the depth difference between any
two adjacent cells is at most one. Then, Lengyel \emph{et al.} propose
specific MC configurations for such transition cells,  leading to a
crack-free triangulation.

@@

\paragraph{Hierarchical Decomposition.}  As discussed above, best
results in terms of texture fetch can be obtained by the approach
presented in Fig.~\ref{fig:approx2}-$b$. At each vertex, the
computation algorithm would consist in a recursive decomposition of
the domain such that a cell is subdivided if it does not belong to
$\Ball{R}{\vx}\cap\Shape$. However, on GPU hardware, this approach is
highly inefficient since the hardware optimizes the parallelism only
when the micro-program described in the shader has a predictive
execution flow. Hence, implementing the recursive algorithm (or its
de-recursified version) requires a lot of branches (conditional
\emph{if} tests) in the flow. Instead of implementing this approach,
we have considered a fast hierarchical approximation following
Fig.~\ref{fig:approx2}-$a$: for a given radius $R$, we precompute a
hierachical decomposition of $B_R$. Such decomposition is made of
octree cells at different resolutions. At a given MC vertex $\vx$, the
algorithm becomes simple since we just scan the ball cells and
estimate the area (or other moments) from the mipmap values associated
to this cell. Note that cells in the $B_R$ decomposition may not be
aligned wuth mipmap cells. However, we can use GPU mipmap
functionalities which interpolates mipmap values if we probe at
non-discrete positions. For a given radius $R$, the expected number of
cells on $B_R$ is in $O(\log{R})$. Implementation details on the
hierarchical octree representation can be found in the supplementary
material (see Section \ref{sec:discussion}).

@@

In this article, we have propose a full data parallel framework on GPU
hardware which combines an adaptive isosurface construction from
digital data with a curvature tensor estimation at each vertex. Using
this approach, we can explore in realtime different curvature
measurements (mean, Gaussian, principal directions) with different
ball radii on potentially large dynamic dataset. Our proposal relies
on both a linear octree representation with Morton codes and an
efficient integral computation on GPU.  The source code and additional
material (video\ldots) are available on the project website
(\url{https://github.com/dcoeurjo/ICTV}).

>>>>

\title{Interactive Curvature Tensor Visualization on Digital
Surfaces\thanks{This work has been mainly funded by \textsc{DigitalSnow}
ANR-11-BS02-009, \textsc{KIDICO} ANR-2010-BLAN-0205 and \textsc{PRIMES Labex}
ANR-11-LABX-0063/ ANR-11-IDEX-0007 research grants.}}

@@

Volumetric objects are being more and more popular in many
applications ranging from object modeling and rendering in Computer
Graphics to geometry processing in Medical Imaging or Material
Sciences. When considering large volumetric data, interactive
visualization of such images (or isovalue surfaces) is a complex
problem. Such issues become even more difficult when dynamic
volumetric datasets are considered. Beside such isosurface
visualization, we are also interested in performing geometry
processing on the digital object and to explore different parameter
settings of the geometry processing tool.  In this work, we focus on
interactive visualization of volumetric data isosurfaces combined with
a curvature tensor (mean/Gaussian curvature, principal curvatures
directions, \ldots) estimation at each point of the digital
surface. Curvature maps on a given isosurface are key information when
exploring 3D structures in Material Sciences \todo{REF}. Furthermore,
most curvature estimators require a parameter fixing the scale
to which the computation is performed. For short, such parameter
(integration radius, convolution kernel size, \ldots) specifies noise
level on the object surface. As a consequence, exploring interactively
different values of such parameter is mandatory.
\vspace{0.2cm}

@@

\sloppy\noindent\textbf{Related works}\quad Extracting and Visualizing
isosurface on volumetric data has been widely investigated since the
seminal Marching Cubes approach by \textsc{Lorensen} and
\textsc{Cline}~\cite{lorensen1987marching}. This approach being data parallel,
GPU implementation of this method is very efficient
\cite{tatarchuk2007real}. However, such technique generates a lot of
triangles which is not well suited to large digital data. Hence,
adaptive approaches have been proposed in order to optimize the
triangular mesh resolution to the object geometry or depending on the
viewpoint. In this topic, many solutions have been developed in
Computer
Graphics~\cite{shu1995adaptive,schaefer2004dual,lengyel2010voxel,DBLP:journals/cgf/LewinerMPPL10,DBLP:journals/cvgip/LobelloDD14}.
The method developed by \textsc{Lengyel} \textit{et al.}
\cite{lengyel2010voxel} is the best suited to our needs.  However, it
was only developed on CPU to visualize in real-time static or near
static data, combining octree with new Marching Cubes configurations
to generate adaptive meshes. In this paper we propose a GPU pipeline
inspired from this method.  It leaves the CPU free for other tasks and
avoids synchronization with the GPU.  Our high framerate also allows
this method to run in real-time on fully dynamic data.  The
triangulation is controlled by a GPU LoD criterion that ensures that
triangles project onto more than one pixel, thus removing geometrical
aliasing.

@@

In \cite{CVIU2014}, authors have also defined the 3D digital principal curvature
estimators $\PrincCurvH{1}{R}$ and $\PrincCurvH{2}{R}$ on $Z\subset\Z^3$ based
on digital moments:
%
\begin{Definition}
  %
  Given the Gauss digitization $Z \EqDef \DSh$ of a shape $\Shape \subset \R^3$,
  {\em 3D digital principal curvature estimators} are defined for any point $\vx
  \in \R^3$ as:
  %
  \begin{align}
    \forall 0 < h < R,\quad
    \PrincCurvH{1}{R}(Z,\vx,h) &\EqDef \frac{6}{\pi {R}^6}(\hat{\lambda}_2 - 3\hat{\lambda}_1) + \frac{8}{5 R} \,,\label{eq:dig-curvature-estimator-k1}\\
    \PrincCurvH{2}{R}(Z,\vx,h) &\EqDef \frac{6}{\pi {R}^6}(\hat{\lambda}_1 - 3\hat{\lambda}_2) + \frac{8}{5 R} \,,
    \label{eq:dig-curvature-estimator-k2}
  \end{align}
  %
  where $\hat{\lambda}_1$ and $\hat{\lambda}_2$ are the two greatest eigenvalues
  of the covariance matrix of the intersection between the digital shape
  $\DigShape$ and a ball of digital radius $R/h$ centered on digital point
  $\vx/h$.
  %
\end{Definition}
%
The covariance matrix needs to compute digital moments of order $0$, order $1$
and order $2$ (see Equation~20 of \cite{CVIU2014} for more details).
% ($\DMom{000}{h}$), of order $1$ ($\DMom{100}{h}$, $\DMom{010}{h}$ and
% $\DMom{001}{h}$) and of order $2$ ($\DMom{200}{h}$, $\DMom{020}{h}$,
% $\DMom{002}{h}$, $\DMom{110}{h}$, $\DMom{011}{h}$ and $\DMom{101}{h}$, see
% Equation~20 of \cite{CVIU2014} for more details).
They have also proved the convergence of these estimators when setting the ball
radius $h$ in $R = kh^\frac{1}{3}$, where $k$ is a constant related to the
maximal curvature of the shape, on convex shapes with at least $C^3$-boundary
and bounded curvature \cite{CVIU2014}. More formally, they get:
%
\begin{align}
  \forall 0 < h \le h_0,\,\, &\forall \vx \in \Bd{\Shape},\,\,
  \forall \vxH \in \Bd{\Body{\DSh}{h}} \text{~with~} \| \vxH -\vx\|_\infty \le h, \nonumber \\
  &\PrincCurvH{1}{R}(\DigF{\Shape}{h},\vxH, h) = \PrincCurv{1}(\Shape,\vx) + O\left(h^{\frac{1}{3}}\right) \,,\\
  &\PrincCurvH{2}{R}(\DigF{\Shape}{h},\vxH, h) = \PrincCurv{2}(\Shape,\vx) + O\left(h^{\frac{1}{3}}\right) \,.
\end{align}

Additionally, eigenvectors associated with $\hat{\lambda}_1$ and
$\hat{\lambda}_2$ of the covariance matrix correspond to principal curvature
directions $\PrincDirH{1}{R}$ and $\PrincDirH{2}{R}$. The smallest eigenvector
corresponds to the normal direction $\NormalDirH{R}$ at $\vx$. Convergence
results can be found in \cite{ChapterIICurvature}.\todo{Changer
ChapterIICurvature dans le bibtex...}

@@

In this section, we detail the adaptive isosurface extraction
algorithm. The proposed approach is based on an octree representation
of the input object on which an efficient adaptive triangulation
of the isosurface is constructed on GPU. Such approach is motivated by
the fact that the hierarchical representation of the object allows us
to handle large datasets and to locally adapt the level of details
\wrt the geometry or camera position. We first present the
octree representation and then the isosurface extraction from
hierarchical octree cells.

@@

\begin{figure}[!htbp]
  \begin{center}
    \subfigure{}{\includegraphics[width=0.9\textwidth]{figs/morton}}
    %%\subfigure{}{\includegraphics[height=0.17\textheight]{figs/partitioning}}
    %%\subfigure{}{\includegraphics[height=0.17\textheight]{figs/quadtree_activefront}}
  \end{center}
\caption{Morton codes associated to cells of a linear quatree. Each
   morton code of a child cell is obtained by adding a suffix to its
  parent code \emph{(left)}. The adaptive representation consists of
  quadtree cells whose depth is view point dependent
  \emph{(middle)}. Finally, adaptive Marching Cubes is used to
  generate the triangulation \emph{(right)}.}
\label{fig_quadtree_partitionning}
\end{figure}

Representing a hierachical structure on GPU is usually challenging since the GPU
memory is limited and non consistent \todo{fix coherent/aligned/coalesced?}
memory access (\emph{e.g.} tree traversal using pointer dereferencing) may lead
to huge payload from cache misses.  Efficient spatial tree encoding can be
achieved using pointerless structures. We use linear quadtrees or octrees to
represent classical quadtree/octree trees, see for instance
\textsc{Gargantini}~\cite{gargantini1982effective}. This structure consists in
indexing each cell by a \emph{Morton code}: the code of children cells are
defined by the code of the parent suffixed by two bits (in dimension 2, three
bits in dimension 3) (see Figure \ref{fig_quadtree_partitionning}-\emph{left}).
A cell's code encodes its position with respect to its parent cell and its
complete path to the tree root. Hence, whatever their depth, all cell codes are
simply stored in a linear vector on memory. Furthermore, cell operations such as
subdivision, merging, neighbors access can be efficiently implemented using
bitwise operations on the Morton code. In the following, we use the GPU
implementation proposed by \textsc{Dupuy} \textit{et al.}~\cite{dupuy2014quadtrees}. By
using their approach, Morton codes and bitwise operations are optimized to match
with GPU hardware requirements.

@@

Using this spatial datastructure, triangulated mesh can be constructed
from a Marching Cubes approach \cite{lorensen1987marching} (MC for short): the
triangulation is generated from local triangle patches computed on
local cell configurations. Since triangle patches are locally
computed, such approach is fully parallel and easy to implement on GPU
hardwares. However, since adjacent cells may not have the same depth
on the octree, original \textsc{Lorensen} and \textsc{Cline}'s rules need to be
updated (see Figure~\ref{fig_quadtree_partitionning}-\emph{right}). Many authors have addressed this problem both for primal and
dual meshes
\cite{shu1995adaptive,schaefer2004dual,lengyel2010voxel,DBLP:journals/cgf/LewinerMPPL10,DBLP:journals/cvgip/LobelloDD14}.

In the following, we use the extension of MC configurations to handle
adaptive structures proposed by \textsc{Lengyel} \emph{et al.}~\cite{lengyel2010voxel}. First this approach constrains the octree
structure in order to make sure that the depth difference between any
two adjacent cells is at most one. Then, \textsc{Lengyel} \emph{et al.} propose
specific MC configurations for such transition cells,  leading to a
crack-free triangulation.

@@

\paragraph{Hierarchical Decomposition.}  As discussed above, best
results in terms of texture fetch can be obtained by the approach
presented in Fig.~\ref{fig:approx2}-$b$. At each vertex, the
computation algorithm would consist in a recursive decomposition of
the domain such that a cell is subdivided if it does not belong to
$\Ball{R}{\vx}\cap\Shape$. However, on GPU hardware, this approach is
highly inefficient since the hardware optimizes the parallelism only
when the micro-program described in the shader has a predictive
execution flow. Hence, implementing the recursive algorithm (or its
de-recursified version) requires a lot of branches (conditional
\emph{if} tests) in the flow. Instead of implementing this approach,
we have considered a fast hierarchical approximation following
Fig.~\ref{fig:approx2}-$a$: for a given radius $R$, we precompute a
hierachical decomposition of $B_R$. Such decomposition is made of
octree cells at different resolutions. At a given MC vertex $\vx$, the
algorithm becomes simple since we just scan the ball cells and
estimate the area (or other moments) from the mipmap values associated
to this cell. Note that cells in the $B_R$ decomposition may not be
aligned with mipmap cells. However, we can use GPU mipmap
functionalities which interpolates mipmap values if we probe at
non-discrete positions. For a given radius $R$, the expected number of
cells on $B_R$ is in $O(\log{R})$. Implementation details on the
hierarchical octree representation can be found in the supplementary
material (see Section \ref{sec:discussion}).

@@

In this article, we have propose a full data parallel framework on GPU
hardware which combines an adaptive isosurface construction from
digital data with a curvature tensor estimation at each vertex. Using
this approach, we can explore in realtime different curvature
measurements (mean, Gaussian, principal directions) with different
ball radii on potentially large dynamic dataset. Our proposal relies
on both a linear octree representation with Morton codes and an
efficient integral computation on GPU.  The source code and additional
material (video, \ldots) are available on the project website
(\url{https://github.com/dcoeurjo/ICTV}).
